### 2014-05-20 text/Chapter04.tex
Adding product and related notions (Chapter 4)

\subsection{Examples}

@@

\begin{example}[Indexed option]
The indexed option comonad is defined over a monoid $(\{ \ident{L},\ident{D} \}, \sqcup,\ident{L})$ 
where $\sqcup$ is defined as earlier, \ie~$\ident{L} = \cclrd{r} \sqcup \cclrd{s} \Longleftrightarrow \cclrd{r}=\cclrd{s}=\ident{L}$.
Assuming $1$ is the unit type inhabited by $()$, the mappings are defined as follows:
\begin{equation*}
\begin{array}{l}
\ctyp{\ident{L}}{\alpha} = \alpha\\
\ctyp{\ident{D}}{\alpha} = 1\\
\\
\ident{counit}_{\ident{L}} : \ctyp{\ident{L}}{\alpha} \rightarrow \alpha \\
\ident{counit}_{\ident{L}}~v = v\\
\end{array}
\qquad
\begin{array}{lcl}
\multicolumn{3}{l}{
  \ident{cobind}_{\cclrd{r},\cclrd{s}} ~:~ (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) 
    \rightarrow (\ctyp{\cclrd{r}\sqcup\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}) }\\
\ident{cobind}_{\ident{L},\ident{L}}~f~x &\narrow{=}& f~x\\
\ident{cobind}_{\ident{L},\ident{D}}~f~() &\narrow{=}& ()\\
\ident{cobind}_{\ident{D},\ident{L}}~f~() &\narrow{=}& f~()\\
\ident{cobind}_{\ident{D},\ident{D}}~f~() &\narrow{=}& ()\\
\end{array}
\end{equation*}
\end{example}

@@

\begin{example}[Indexed product]
The semantics of implicit parameters

\end{example}

\paragraph{Related work}

Not a family of comonads

Katsumata

@@

\paragraph{Example}
Indexed comonad are analogous to comonads (in coKleisli form), but
with the additional monoidal structure on indices.  Indeed, comonads
are a special case of indexed comonads with a trivial singleton
monoid, \eg{}, $(\{1\}, \ast, 1)$ with $1 \ast 1 = 1$ where $\ccat^1$
is the underlying functor of the comonad and $\counit$ and
$\cobind{(-)}{1, 1}$ are the usual comonad operations.
However, as demonstrated next, not all indexed comonads are derived from ordinary comonads.

\paragraph{Example}
The \emph{indexed partiality comonad} encodes free-variable contexts
of a computation which are either \emph{live} or \emph{dead} (\ie{}, have
\emph{liveness} coeffects) with the monoid $(\{\ident{D}, \ident{L}\},
\sqcap, \ident{L})$, where $\ccat^{\ident L} A = A$ encodes live
contexts and $\ccat^{\ident D} A = 1$ encodes dead contexts,
where $1$ is the unit type inhabited by a single value $()$.
The \emph{counit} operation $\counit{} : \ccat^{\ident{L}} A \rightarrow A$
is defined $\counit\ x = x$ and \emph{coextend},
for all $f : \ccat^r A \rightarrow B$, and thus
$\cobind{f}{r, s} : \; \ccat^{r \sqcap s} A \rightarrow \ccat^s B$,
is defined:
%
\begin{align*}
\begin{array}{llll}
\cobind{f}{\ident{D}, \ident{D}} x = () \qquad\qquad & 
\cobind{f}{\ident{D}, \ident{L}} x = f () \qquad\qquad &
\cobind{f}{\ident{L}, \ident{D}} x = () \qquad\qquad &
\cobind{f}{\ident{L}, \ident{L}} x = f~x 
\end{array}
\end{align*}
%
The indexed family $C^r$ here is analogous to the non-indexed \textsf{Maybe} (or \emph{option})
data type $\mathsf{Maybe}\ A = A + 1$.  This
type does not permit a comonad structure since
$\varepsilon : \mathsf{Maybe}\ A \rightarrow A$ is undefined at $(\textsf{inr} \, ())$. 
For the indexed comonad, $\counit$ need only be defined
 for $\ccat^{\ident{L}} A = A$. Thus, indexed comonads capture a broader range of
 contextual notions of computation than comonads.

 Moreover, indexed comonads are not restricted by the \emph{shape
   preservation} property of comonads~\cite{orchard12codo}: that
 a coextended function cannot change the \emph{shape} of the
 context. For example, in the second case above $f^\dagger_{\ident{D},
   \ident{L}} : \ccat^{\ident{D}} A \rightarrow \ccat^{\ident{L}} B$
 where the shape changes from $1$ (empty context) to $B$ (available
 context). 
a

\subsection{Indexed monoidal comonads}

!!!

not a family of comonads

@@

\section{Coeffect semantics using indexed comonads}
\label{sec:comonads}

\vspace{-1em}
\subsection{Monoidal indexed comonads.}
Indexed comonads provide a semantics to sequential composition, but
additional structure is needed for the semantics of the full coeffect calculus.
Uustalu and Vene~\cite{comonads-notions} additionally require a (\emph{lax semi-}) \emph{monoidal comonad} 
structure, which provides a monoidal operation $\mathsf{m} : C A
\times C B \rightarrow C (A \times B)$ for merging contexts (used in
the semantics of abstraction).

>>>>

\subsection{Examples of indexed comonads}

@@

\begin{example}[Indexed option]
The indexed option comonad is defined over a monoid $(\{ \ident{L},\ident{D} \}, \sqcup,\ident{L})$ 
where $\sqcup$ is defined as earlier, \ie~$\ident{L} = \cclrd{r} \sqcup \cclrd{s} \Longleftrightarrow \cclrd{r}=\cclrd{s}=\ident{L}$.
Assuming $1$ is the unit type inhabited by $()$, the mappings are defined as follows:
%
\begin{equation*}
\begin{array}{l}
\ctyp{\ident{L}}{\alpha} = \alpha\\
\ctyp{\ident{D}}{\alpha} = 1\\
\\
\ident{counit}_{\ident{L}} : \ctyp{\ident{L}}{\alpha} \rightarrow \alpha \\
\ident{counit}_{\ident{L}}~v = v\\
\end{array}
\qquad
\begin{array}{lcl}
\multicolumn{3}{l}{
  \ident{cobind}_{\cclrd{r},\cclrd{s}} ~:~ (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) 
    \rightarrow (\ctyp{\cclrd{r}\sqcup\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}) }\\
\ident{cobind}_{\ident{L},\ident{L}}~f~x &\narrow{=}& f~x\\
\ident{cobind}_{\ident{L},\ident{D}}~f~() &\narrow{=}& ()\\
\ident{cobind}_{\ident{D},\ident{L}}~f~() &\narrow{=}& f~()\\
\ident{cobind}_{\ident{D},\ident{D}}~f~() &\narrow{=}& ()\\
\end{array}
\end{equation*}
\end{example}

@@

\begin{example}[Indexed product]
The semantics of implicit parameters is modelled by an indexed product comonad. We use a monoid
$(\mathcal{P}(\ident{Id}), \cup, \emptyset)$ where \ident{Id} is the set of (implicit parameter) names.
As previously, all parameters have the type $\rho$. The data type $\ctyp{\cclrd{r}}{\alpha}
= \alpha \times (\cclrd{r} \rightarrow R)$ represents a value $\alpha$ together with a function that
associates a parameter value $\rho$ with every implicit parameter name in $\cclrd{r} \subseteq \ident{Id}$.
The cobind and counit operations are defined as:
%
\begin{equation*}
\begin{array}{l}
\ident{counit}_{\emptyset} : \ctyp{\emptyset}{\alpha} \rightarrow \alpha \\[-0.25em]
\ident{counit}_{\emptyset}~(a, g) = a\\[0.5em]
\end{array}
\quad
\begin{array}{l}
\ident{cobind}_{\cclrd{r},\cclrd{s}} ~:~ (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) 
    \rightarrow (\ctyp{\cclrd{r}\cup\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta})\\[-0.25em]
\ident{cobind}_{\cclrd{r},\cclrd{s}}~f~(a,g) = (f(a,\restr{g}{\cclrd{r}}), \restr{g}{\cclrd{s}})\\[0.5em]
\end{array}
\end{equation*}
\end{example}

\noindent
The definition of \ident{counit} simply ignores the function and returns the value in the context.
The \ident{cobind} operation uses the restriction operation $\restr{f}{\cclrd{r}}$, which we 
already defined when discussing semantics of implicit parameters in Section~\ref{sec:applications-flat-impl}
(indeed, \ident{cobind} here captures an essential part of the semantics).

The function $g$ in \ident{cobind} is defined on the union of the implicit parameters, 
\ie~$\cclrd{r}\cup\cclrd{s}\rightarrow\rho$. When passing it to $f$, we restrict it to
just $\cclrd{r}$ and when returning it as a result, we restrict it to $\cclrd{s}$.

\subsection{Properties and related notions}

We discuss additional examples in Section~\ref{sec:flat-semantics-monoidal}, after we look at the
remaining structure that is needed to define the semantics of flat coeffect calculus. Before doing 
so, we discuss additional properties and categorical structures that have been proposed mainly in 
the context of monads and effects and are related to indexed comonads.

\paragraph{Shape preservation.}
Ordinary comonads have the \emph{shape preservation} property \cite{comonads-codo}. Intuitively, 
this means that the shape of the additional context does not change during the computation. For
example, in the \ident{NEList} comonad, the length of the list stays the same after applying 
\ident{cobind}.

Indexed comonads are not restricted by this property of comonads. For example, given the indexed
product monad, in the computation $\ident{cobind}_{\cclrd{r}, \cclrd{s}} f$ above, the shape of
the context changes from containing implicit parameters $\cclrd{r} \cup \cclrd{s}$  to containing
just implicit parameters $\cclrd{s}$.

\paragraph{Families of monads.}
When linking effect systems and monads, Wadler and Thiemann \cite{monad-notions} propose a
\emph{family of monads} as the categorical structure. The dual structure, \emph{family of
comonads}, is defined as follows.

\begin{definition}
A \emph{family of comonads} is formed by triples $(\ctyp{\cclrd{r}}{}, \ident{cobind}_{\cclrd{r}}, 
  \ident{counit}_{\cclrd{r}})$ for all $\cclrd{r}$ such that each triple forms a comonad. Given 
$\cclrd{r}, \cclrd{r'}$ such that $\cclrd{r} \leq \cclrd{r'}$, there is also a mapping 
$\iota_{\cclrd{r'}, \cclrd{r}} : \ctyp{\cclrd{r'}}{} \rightarrow \ctyp{\cclrd{r}}{}$ satisfying
certain coherence conditions.
\end{definition}

Family of comonads is more restrictive than indexed comonad, because each of the data types needs
to form a comonad separately. For example, our indexed option does not form a family of comonads
(again, because \ident{counit} is not defined on $\ctyp{\ident{D}}{\alpha}=1$). However, given a 
family of comonads and indices such that $\cclrd{r} \leq \cclrd{r}\cseq\cclrd{s}$, we can define 
an indexed comonad. Briefly, to define $\ident{cobind}_{\cclrd{r},\cclrd{s}}$ of an indexed comonad, 
we use $\ident{cobind}_{\cclrd{r}\cseq\cclrd{s}}$ from the family, together with two lifting operations:
$\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{r}}$ and $\iota_{\cclrd{r}\cseq\cclrd{s}, \cclrd{s}}$.

\paragraph{Parameteric effect monads.}
Parametric effect monads introduced by Katsumata \cite{monads-parametric} (independently to our 
indexed comonads) are closely related to our definition.  Although presented in a more general 
categorical framework (and using monads), the model defines \ident{unit} operation only on the 
unit of a monoid and \ident{bind} operation composes effect annotations using the provided monoidal 
structure.

\subsection{Indexed monoidal comonads}
\label{sec:flat-semantics-monoidal}

@@

Indexed comonads provide a semantics to sequential composition, but
additional structure is needed for the semantics of the full coeffect calculus.
Uustalu and Vene~\cite{comonads-notions} additionally require a (\emph{lax semi-}) \emph{monoidal comonad} 
structure, which provides a monoidal operation $\mathsf{m} : C A
\times C B \rightarrow C (A \times B)$ for merging contexts (used in
the semantics of abstraction).

@@
